---
title: "##ECON 294A Final Exam"
author: "Dylan McCombs"
date: "March 14, 2016"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warnings = F)
```

### Introduction
Upon first reading the assignment handout, I was excited to get started. I figured that the work that it involved was going to be quite challenging, given the fact that my classmates and I had yet to actually use sqlite and many of the tools that it entails. However, once I had started this assignment, I realized that the challenge of obtaining decent and useful results was even greater than initially thought. 

Finding the relationship of departure delays and flight cancellations with the following categories proved quite difficult:

a) weather

b) time of day, day of week, time of year, and other time aspects

c) airport destination

d) plane charactersitics

Unfortunately, the results shown in my "Analysis" section are scarce and unhelpful, and I could not figure out why I was getting an error message stating that the flights_weather data was not found as I was able to run it in R before putting the code in this .Rmd file.

```{r,echo=F}
library("dplyr")  #for dplyr
# install.packages("RSQLite")
library("RSQLite") #for sqllite
# install.packages("nycflights13")
library("nycflights13")

###### Obtaining & organizing data ######

### Creating a new empty database ###
my_294A.db <- src_sqlite("my_294A.db.sqlite3", create = T)
my_294A.db



```{r, echo=F, eval=F}
### Creating df tables in my db ###
weather_sqlite <- copy_to(
  my_294A.db, weather, temporary = FALSE, 
  indexes = list(
    c("year", "month","day","hour"),
    "origin")
)

planes_sqlite <- copy_to(
  my_294A.db, planes, temporary = FALSE, 
  indexes = list("tailnum", "engine", "type", "seats")
)

flights_sqlite <- copy_to(
  my_294A.db, flights, temporary = FALSE, 
  indexes = list(
    c("year", "month", "day"), "dest", "origin", 
    "dep_delay",  "tailnum", "dep_time", "arr_time")
)

airlines_sqlite <- copy_to(
  my_294A.db, airlines, temporary = FALSE, 
  indexes = list("carrier")
)

airports_sqlite <- copy_to(
  my_294A.db, airports, temporary = FALSE, 
  indexes = list("faa", "lat", "lon", "alt")
)

# caching data
nycflights13_sqlite()
my_294A.db

# Creating dataframe tables from src query:
flights_sqlite <- tbl(nycflights13_sqlite(), "flights")
flights_sqlite
#airports_sqlite <- tbl(nycflights13_sqlite(), "airports")
#airports_sqlite
planes_sqlite <- tbl(nycflights13_sqlite(), "planes")
planes_sqlite
weather_sqlite <- tbl(nycflights13_sqlite(), "weather")
weather_sqlite

```

```{r, echo=F, eval=T}

# merging flights and planes by tailnum
# merging flights and weather by common names

# naturally joining flights & weather by: 
# c("year", "month", "day", "origin", "hour")
flights_weather <- inner_join(flights, weather) #117251 obs.

# creating other vars. & organizing dataframe
flights_weather <- flights_weather %>%
  mutate(
    date           = paste(year, month, day, sep = "-"),
    date           = as.Date(date, "%Y-%m-%d"),
    cancellations  = ifelse(is.na(arr_time), 1, 0),
    sig_dep_delays = ifelse(dep_delay>30, 1, 0),
    dest           = as.factor(dest),
    carrier        = as.factor(carrier),
    dep_time       = as.ts(dep_time),
    month          = as.factor(month)
  ) %>%
  select(-year, -day) %>%
  select(date, dep_delay, arr_delay, cancellations, 
         sig_dep_delays, everything())

sum(flights_weather$cancellations==1, na.rm=T) # 151 cancellations
sum(flights_weather$sig_dep_delays==1, na.rm=T) # 19873 sig. dep. delays


# naturally joining flights & planes by: 
# c("year", "tailnum")
flights_planes <- inner_join(flights, planes, 
                             by="tailnum") #284170 obs.

# creating other vars. & organizing dataframe
flights_planes <- flights_planes %>% 
  rename(year = year.x, year_built = year.y) %>%
  mutate(
    date           = paste(year, month, day, sep = "-"),
    date           = as.Date(date, "%Y-%m-%d"),
    cancellations  = ifelse(is.na(arr_time),1,0),
    sig_dep_delays = ifelse(dep_delay>30, 1, 0),
    dest           = as.factor(dest),
    carrier        = as.factor(carrier),
    dep_time       = as.ts(dep_time),
    month          = as.factor(month),
    year_built     = as.factor(year_built),
    type           = as.factor(type),
    manufacturer   = as.factor(manufacturer),
    model          = as.factor(model),
    engines        = as.factor(engines),
    seats          = as.factor(seats),
    engine_type    = as.factor(engine)
  ) %>%
  select(-year, -day) %>%
  select(date, dep_delay, arr_delay, cancellations, 
         sig_dep_delays, everything())

sum(flights_planes$cancellations==1, na.rm=T) # 4551 obs.
sum(flights_planes$sig_dep_delays==1, na.rm=T) # 42067 obs.

###### Simple Analyses ######
library(ggplot2)

```

### Analyses

```{r, echo=F, eval=T}
FW.glm1 <- glm(sig_dep_delays~temp+dewp+humid+wind_speed+
                 precip+pressure+visib, flights_weather,
                 family=binomial(link='logit'))
```

```{r, echo=F}
summary(FW.glm1)
```

```{r, echo=F, eval=T}
FW.glm2 <- glm(cancellations~temp+dewp+humid+wind_speed+
                 precip+pressure+visib, flights_weather, 
                 family=binomial(link='logit'))
```

```{r, echo=F, eval=T}
summary(FW.glm2)
```

### Conclusions








